---
title: "Data manipulation with dplyr"
output: github_document
---

Once you've imported data, you're going to need to do some cleaning up

#CSV Data Import + Cleaning
```{r, echo=FALSE, message=FALSE}
library(tidyverse)

options(tibble.print_min = 3)

litters_data = read_csv("./data_import_examples/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data_import_examples/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)
```

#Select columns: `select`
Select columns in your data frame
```{r}
select(litters_data,
       group, litter_number, gd0_weight, pups_born_alive) #4 columns selected (method1)

select(litters_data, group:gd_of_birth) #4 columns selected (method2)
```

Remove several columns in your data frame
```{r}
select(litters_data, -pups_survive)
```

`Rename` some variables
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for select; read about all of them using `?select_helpers`. 
* I use `starts_with()`, `ends_with()`, and `contains()` often, especially when there variables are named with suffixes or other standard patterns
* I also frequently use is `everything()`, which is handy for reorganizing columns without discarding anything
* `relocate` does a similar thing (and is sort of like rename in that it’s handy but not critical)
```{r}
select(litters_data, starts_with("gd"), ends_with("weight"))

#changing the order by everything ()
#i.e., take everything else to stick at the end of the 2 selected variables
select(litters_data, litter_number, pups_survive, everything())

#relocate
relocate(litters_data, litter_number, pups_survive)
```

In larger datasets,

Lastly, like other functions in `dplyr`, select will export a dataframe even if you only select one column. Mostly this is fine, but sometimes you want the vector stored in the column. To pull a single variable, use `pull`.



# Selecting rows: `filter`
You will often filter using comparison operators (`>`, `>=`, `<`, `<=`, `==`, and `!=`). You may also use `%in%` to detect if values appear in a set, and `is.na()` to find missing values. The results of comparisons are logical – the statement is `TRUE` or `FALSE` depending on the values you compare – and can be combined with other comparisons using the logical operators `&` and `|`, or negated using `!`.
```{r}
filter(litters_data,
          gd_of_birth == 20,
          pups_born_alive >= 2,
          pups_survive != 4,
          !(pups_survive == 4),
          group %in% c("Con7", "Con8"),
          group == "Con7" & gd_of_birth == 20)
```


A very common filtering step requires you to omit missing observations. You can do this with `filter`, but I recommend using `drop_na` from the `tidyr` package:
* `drop_na(litters_data)` will remove ANY ROW with a missing value
* `drop_na(litters_data, wt_increase)` will remove rows for which `wt_increase` is missing.

#Change various variables at one time: `Mutate`
Sometimes you need to select columns; sometimes you need to change them or create new ones. You can do this using `mutate`.

The example below creates a new variable measuring the difference between `gd18_weight` and `gd0_weight` and modifies the existing group variable.
```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight, #calculate the difference of weights
  group = str_to_lower(group) #change all the string to lower cases
)
```

A few things in this example are worth noting:

* Your new variables can be functions of old variables
* New variables appear at the end of the dataset in the order that they are created
* You can overwrite old variables
* You can create a new variable and immediately refer to (or change) it


#Sort the variable: `arrange`
You can arrange the rows in your data according to the values in one or more columns:
```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10) 
  #desc() means to sort descending
  #10 is to show the top 10 rows
```

#Pipe: `%>%`
For example, suppose you want to load the data, clean the column names, remove `pups_survive`, and create `wt_gain`. There are a couple of options for this kind of multi-step data manipulation:

define intermediate datasets (or overwrite data at each stage)
* nest function calls
* The following is an example of the pipe option:
```{r}
litters_data = 
  read_csv("./data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)

litters_data
```

