---
title: "Data import"
output: github_document
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(haven)
```

##Data import: CSVs

Let's import data using the 'readr' package.

```{r}
#using a relative path is more expandable when transfer into another computer (compared with the absolute path)!#
litters_df = read_csv("data_import_examples/FAS_litters.csv")
litters_df = janitor::clean_names(litters_df) #clean up the package functions#
```

Look at the data
```{r}
litters_df 

head(litters_df)
tail(litters_df)
```

```{r, eval = FALSE}
view(litters_df)
```

summary of the data
```{r}
skimr::skim(litters_df)
```

`read_csv` options ..
```{r}
#customize the data treated as missing value#
read_csv("data_import_examples/FAS_litters.csv", na = c("","NA",999,88))
#if you want to skip the first two rolls which works as notes
read_csv("data_import_examples/FAS_litters.csv", skip = 2)
```

##other file formats like excel
We need to read in an excel spreadsheet...
```{r}
mlb_df = read_excel("data_import_examples/mlb11.xlsx")
```

```{r,eval=FALSE}
view(mlb_df)
```

Selectively read a specific area in an excel file
```{r}
lotr_words = 
  read_excel(
    "data_import_examples/LotR_Words.xlsx",
    range = "B3:D6")
```

##Still more formats (SAS file)..
Read in a SAS dataset
```{r}
#Haven package required#
pulse_df = read_sas("data_import_examples/public_pulse_data.sas7bdat")
```

##Data export
```{r}
write_csv(lotr_words, file = "data_import_examples/lotr_words.csv")
```

##Why not base R???
```{r}
dont_do_this_df = read.csv("data_import_examples/FAS_litters.csv")
```
Less organized!!
Thus, we prefer "read_csv" in readr package, rather than "read/csv" in base R.



